Environmental and geographic variables are effective surrogates for genetic variation in conservation planning

### 2017-10-18 code/rmarkdown/article.Rmd
minor tweaks to ms changes to analysis that were not pushed clean up documentation

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Earth and Environmental Sciences, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  JOH performed the analyses. All authors developed the study and drafted the manuscript.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for the processes that generate and maintain biodiversity to continue. However, genetic variation, the raw material for evolution, is difficult to capture in conservation planning, not least because genetic data require considerable resources to obtain and analyze. Here we show that freely available environmental and geographic distance variables are highly effective surrogates in conservation planning for representing adaptive and neutral intra-specific genetic variation. We obtained occurrence and genetic data from the IntraBioDiv project for 27 alpine plant species collected over the European Alps using a gridded sampling scheme. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the grid cells. We then used the cells as planning units to prioritize protected area acquisitions. First, we verified that the spatial patterns of environmental and geographic variation were correlated, respectively, with adaptive and neutral genetic variation. Second, we showed that these surrogates can predict the proportion of genetic variation secured in randomly generated solutions. Finally, we discovered that solutions based only on surrogate information secured substantial amounts of adaptive and neutral genetic variation. Our work paves the way for widespread integration of surrogates for genetic variation into conservation planning.
significance: |
  To protect biodiversity for the long term, nature reserves and other protected areas need to represent a broad range of different genetic types. However, genetic data are expensive and time consuming to obtain. Here we show that freely available environmental and geographic variables can be used as effective surrogates for genetic data in conservation planning. This means that conservation planners can, with some confidence, design protected area systems to represent intra-specific genetic diversity without investing in expensive programmes to obtain and analyze genetic data.
acknowledgements: |
  JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship.
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

<!-- Introduction -->
\dropcap{P}rotected areas spearhead conservation efforts ([@r507]). They buffer species from anthropogenic impacts, providing places for them to persist. Since the resources available for conservation are limited, protected areas need to be sited in places that fulfill conservation objectives for minimal cost ([@r255]). To achieve this, conservation planning exercises often generate plans for entire networks of protected areas (prioritizations) to allocate resources cost-effectively and preserve broad-scale biodiversity processes ([@r9]). Intra-specific genetic variation is fundamental to long-term species persistence ([@r165; @r504]), and as a consequence, there has been increasing interest in designing prioritizations that represent this aspect of biodiversity ([@r67; @r3; @r499; @r61; @r503]).

Although the strength of natural selection is a continuous force, genetic variation is often classified as adaptive or neutral (reviewed in [@r480]). Adaptive genetic variation is associated with loci that significantly affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals (e.g. [@r456; @r487]), though phenotypic-genetic associations are more likely to identify functionally important genomic regions for individual species (e.g. [@r500]). By representing the full range of adaptive variation, protected areas can fortify a species' capacity to persist in a range of different conditions ([@r3; @r43]). In contrast, neutral genetic variation is associated with loci that do not significantly affect fitness, but instead reflect the evolutionary history of different populations. By representing the full range of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity ([@r67]). Thus optimally sited protected area networks would aim to represent patterns of both adaptive and neutral genetic variation ([@r67; @r3; @r43]).

Recently, the use of surrogate data has been proposed to generate prioritizations that capture intra-specific genetic variation ([@r2; @r506; @r200])---without needing to utilize genetic data to achieve this. Since adaptation is ultimately driven by selection pressures, environmental variables have been proposed as surrogates for adaptive genetic variation (e.g. [@r202; @r2]). Specifically, by conserving individuals in a broad range of environmental conditions, we might expect to capture individuals with a diverse set of local adaptations, and overall capture a large proportion of the adaptive genetic variation present in the species. However, the effectiveness of this approach remains unverified. Neutral genetic variation arises from a reduction in gene flow between populations. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood (reviewed in [@r501]). Potential surrogates for neutral genetic variation have been based on variables that predict the level of connectivity between different areas. For instance, according the isolation by distance model (IBD; [@r259]), populations located further apart are predicted to experience less gene flow, and in turn, share fewer neutral loci. Based on this idea, spreading conservation priorities evenly across the geographic distribution of an island network has been found to capture more neutral genetic variation ([@r200]). However this remains untested in spatially contiguous systems where connectivity is complicated by many additional factors (e.g. land use; [@r497])---the typical situation in most conservation planning exercises.

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods---that solely used genetic data and did not use environmental data---to identify loci under selection. These methods returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD; Table S1). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions was needed to sufficiently describe their patterns of adaptive ($K = `r unique(na.omit(adaptive.k))`$ for all species; Table S2) and neutral genetic variation ($`r min(neutral.k)` \leq K \leq `r max(neutral.k)`$ for all species; Table S1). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figures S1--S27) generally showed substantial spatial auto-correlation (e.g. _Cerastium uniflorum_, Figure S34; _Dryas octopetala_, Figure S36), suggesting that the environmental and geographic variables have potential to be effective surrogates for genetic variation.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental variation correlated with the broad-scale patterns in adaptive genetic variation for each species. We also verified that there was a correlation between neutral genetic variation and variation in geographic position. For each species, we constructed dissimilarity matrices expressing differences between the planning units where individuals were detected based on the units' (i) environmental characteristics, (ii) geographic position, and also the (iii) adaptive (if adaptive loci detected), and (iv) neutral genetic characteristics of the individuals found inside them. The spatial patterns of environmental variation were significantly correlated with the pattern of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models; Table S3). Similarly, geographic distance between planning units was also correlated significantly with the spatial pattern of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models; Table S3). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these patterns were correlated, our next step was to determine whether these associations were strong enough to be effective for conservation planning.

<!-- surrogacy analysis -->
For each species, we generated a suite of `r format(raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates, big.mark=',', scientific=FALSE)` random prioritizations, and calculated the proportion of environmental variation and adaptive (if present) genetic variation that each prioritization captured. We then repeated this process, and calculated the proportion of variation in geographic position and neutral genetic variation they sampled. The environmental and geographic variables were moderately effective predictors for the genetic variation represented by the randomly generated prioritizations (Figures 1, S28 and S29). The relationship between the proportion of genetic and surrogate variation secured in the prioritizations varied among the species (species $\times$ proportion of environmental variation interaction term: ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$; species $\times$ proportion of geographic variation interaction term: ${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P =$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that these relationships were positive for all species (environmental: minimum \textit{z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S4). After establishing that the the proportion of surrogate variation secured in a prioritization also predicted the proportion of genetic variation it secured, our final step was to determine if prioritizations were more effective when generated using targets based on the surrogate variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4.5, out.width='3.42in', out.height='2.5in', fig.cap=paste0('The proportion of adaptive (a; ',sum(!is.na(correlation.model.summary.DF$Environmental.corr.P)),' species) and neutral (b; ',sum(!is.na(correlation.model.summary.DF$Geographic.corr.P)),' species) genetic variation represented in a suite of randomly generated prioritizations for different species as a function of the surrogate variation they also secured (lines). Panels (c) and (d) show the distribution of $R^2$ values for the ',sum(!is.na(correlation.model.summary.DF$Environmental.corr.P)) + sum(!is.na(correlation.model.summary.DF$Geographic.corr.P)),' models shown in panels (a) and (b).')}
# make figure
p1 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=env.surrogacy.DF) +
  xlab('Environmental variation\nsecured (%)') +
  ylab('Adaptive genetic\nvariation secured (%)') +
  theme_classic() + theme(axis.line.x=element_line(color='black'),
    axis.line.y=element_line(color='black')) +
  coord_cartesian(xlim=c(0, 1)) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_y_continuous(limits=c(0,1), labels=function(x) {x*100})
for (i in seq_along(env.surrogacy.LST))
  p1 <- p1 + stat_smooth(method='glm', method.args=list(family='binomial'),
    data=filter(env.surrogacy.DF, Species==unique(Species)[i]),
    fullrange=TRUE, se=FALSE, colour='black')
p1 <- p1 + geom_text(x=-Inf, y=Inf, hjust=-0.25, vjust=1.5, label='(a)')

p2 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=geo.surrogacy.DF) +
  xlab('Geographic variation\nsecured (%)') +
  ylab('Neutral genetic\nvariation secured (%)') +
  theme_classic() + theme(axis.line.x=element_line(color='black'),
    axis.line.y=element_line(color='black', colour='transparent')) +
  coord_cartesian(xlim=c(0, 1)) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_y_continuous(limits=c(0,1), labels=function(x) {x*100})
for (i in seq_along(geo.surrogacy.LST))
  p2 <- p2 + stat_smooth(method='glm', method.args=list(family='binomial'),
    data=filter(geo.surrogacy.DF, Species==unique(Species)[i]),
    fullrange=TRUE, se=FALSE, colour='black')
p2 <- p2 + geom_text(x=-Inf, y=Inf, hjust=-0.25, vjust=1.5, label='(b)')

@@

<!-- scenario analysis -->
To determine whether environmental and geographic targets could improve the effectiveness of prioritizations in representing genetic variation, we generated prioritizations using (i) "amount targets" reflecting the traditional approach of representing a certain proportion of each species' geographic distribution, (ii) "amount and surrogate targets" where we targeted representation of the environmental and geographic surrogates as well as a certain proportion of each species' geographic distribution, and (iii) "amount and genetic targets" where we targeted representation of the directly measured genetic variation as well as a certain proportion of each species' geographic distribution. To determine if these patterns were robust, for each of the three combinations of the targets, we generated prioritizations under four scenarios: (i) single-species with equal costs, (ii) single species with acquisition costs, (iii) multi-species with equal costs, and (iv) multi-species with acquisition costs (Figure S30).

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=(5*0.9), fig.height=(7.1*0.9), out.width='3.42in', out.height='4in', fig.cap='The proportion of adaptive and neutral genetic variation secured for each species in solutions based on different targets and scenarios. Box plots show the median, and the 25th and 75th percentiles. Whiskers show the data at 1.5 times the inter-quantile range. Points show data outside the whiskers. Red lines show the target proportion of genetic variation.'}
# prepare data for plotting
scenario.SDF <- scenario.DF %>%
  group_by(Prioritisation, Context, Metric, Species) %>%
  summarize(genetic.held=mean(genetic.held, na.rm=TRUE)) %>%
  data.frame()

@@

The proportion of genetic variation secured in a prioritization depended on the targets used to generate the prioritization (${\chi^2}_{`r scenario.ANOVA.4.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.4.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.4.DF$p.value[[2]])`)$. Most notably, prioritizations based on the surrogate variables represented significantly more genetic variation than amount-based prioritizations (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` SD overall genetic variation secured versus `r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[1]],2)`$; $P `r pvalString(scenario.GLHT.2.DF$p.value[[1]])`$), and were not distinguishable from prioritizations based on measured genetic data (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[3]],2)`$; $P `r pvalString(scenario.GLHT.2.DF$p.value[[3]])`$). These results indicate that environmental and geographic surrogates in this case far outperform traditional amount-based conservation planning, and perform almost as well as a conservation plan based directly on measured genetic data.

>>>>

---
title: "Environmental and geographic variables are effective surrogates for genetic variation in conservation planning"
author:
  - name: Jeffrey O. Hanson
    affiliation: a,1
  - name: Jonathan R. Rhodes
    affiliation: b
  - name: Cynthia Riginos
    affiliation: a
  - name: Richard A. Fuller
    affiliation: a
address:
  - code: a
    address: School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  - code: b
    address: School of Earth and Environmental Sciences, The University of Queensland, Brisbane, QLD, Australia
corresponding_author:
  - code: 1
  - text: "To whom correspondence should be addressed. E-mail: jeffrey.hanson@uqconnect.edu.au"
author_contributions: |
  JOH performed the analyses. All authors developed the study and drafted the manuscript.
lead_author_surname: "Hanson"
abstract: |
  Protected areas buffer species from anthropogenic threats and provide places for the processes that generate and maintain biodiversity to continue. However, genetic variation, the raw material for evolution, is difficult to capture in conservation planning, not least because genetic data require considerable resources to obtain and analyze. Here we show that freely available environmental and geographic distance variables can be highly effective surrogates in conservation planning for representing adaptive and neutral intra-specific genetic variation. We obtained occurrence and genetic data from the IntraBioDiv project for 27 plant species collected over the European Alps using a gridded sampling scheme. For each species, we identified loci that were potentially under selection using outlier loci methods, and mapped their main gradients of adaptive and neutral genetic variation across the grid cells. We then used the cells as planning units to prioritize protected area acquisitions. First, we verified that the spatial patterns of environmental and geographic variation were correlated, respectively, with adaptive and neutral genetic variation. Second, we showed that these surrogates can predict the proportion of genetic variation secured in randomly generated solutions. Finally, we discovered that solutions based only on surrogate information secured substantial amounts of adaptive and neutral genetic variation. Our work paves the way for widespread integration of surrogates for genetic variation into conservation planning.
significance: |
  To protect biodiversity for the long term, nature reserves and other protected areas need to represent a broad range of different genetic types. However, genetic data are expensive and time consuming to obtain. Here we show that freely available environmental and geographic variables can be used as effective surrogates for genetic data in conservation planning. This means that conservation planners can, with some confidence, design protected area systems to represent intra-specific genetic diversity without investing in expensive programmes to obtain and analyze genetic data.
acknowledgements: |
  JOH is supported by an Australian Government Research Training Program (RTP) Scholarship. RAF has an Australian Research Council Future Fellowship.
keywords:
  - conservation
  - biodiversity
  - AFLP
  - genetic diversity
  - alpine plants
pnas_type: pnasresearcharticle
bibliography: references.bib
csl: reference-style.csl
output:
  rticles::pnas_article:
    includes:
      in_header: [preamble.tex]
---

@@

<!-- Introduction -->
\dropcap{P}rotected areas spearhead conservation efforts ([@r507]). They buffer species from anthropogenic impacts, providing places for them to persist. Since the resources available for conservation are limited, protected areas need to be sited in places that fulfill conservation objectives for minimal cost ([@r255]). To achieve this, conservation planning exercises often generate plans for entire networks of protected areas (prioritizations) to preserve species and broad-scale biodiversity processes ([@r9]). After determining which species to protect (e.g. using evolutionary distinctiveness; [@r508]), conservation planners need to ensure that prioritizations represent the intra-specific genetic variation found within species to secure their long-term persistence ([@r3; @r63; @r61]). As a consequence, there has been increasing interest in designing prioritizations that fulfill this objective ([@r368; @r503; @r506; @r515; @r465]).

Although the strength of natural selection is a continuous force, genetic variation is often classified as adaptive or neutral (reviewed in [@r480]). Adaptive genetic variation is associated with loci that significantly affect fitness. Typically, such "adaptive" loci are detected due to anomalous patterns among individuals (e.g. [@r456; @r487]), though phenotypic-genetic associations are more likely to identify functionally important genomic regions for individual species (e.g. [@r500]). By representing the full range of adaptive variation, including adaptations of which we are currently unaware, protected areas can enhance a species' capacity to persist in a range of different conditions ([@r3; @r43]). In contrast, neutral genetic variation is associated with loci that do not significantly affect fitness, but instead reflect the evolutionary history of different populations. By representing the full range of neutral variation, protected areas can safeguard against the adverse effects of low genetic diversity ([@r5]). Thus optimally sited protected area networks would represent patterns of both adaptive and neutral genetic variation ([@r63]).

Recently, the use of surrogate data has been proposed to generate prioritizations that capture intra-specific genetic variation ([@r2; @r506; @r200])---without needing to utilize genetic data directly. Since adaptation is ultimately driven by selection pressures, environmental variables have been proposed as surrogates for adaptive genetic variation (e.g. [@r9; @r2]). Specifically, by conserving individuals in a broad range of environmental conditions, we might expect to capture individuals with a diverse set of local adaptations, and overall capture a large proportion of the adaptive genetic variation present in the species. However, the effectiveness of this approach remains unverified. Neutral genetic variation arises from a reduction in gene flow between populations. Over the last few years, the field of phylogeography has predominantly focused on describing neutral genetic diversity, and so the landscape factors that affect neutral variation are relatively well understood (reviewed in [@r501]). Potential surrogates for neutral genetic variation have been based on variables that predict the level of connectivity between different areas. For instance, according the isolation by distance model (IBD; [@r259]), populations located further apart are predicted to experience less gene flow, and in turn, share fewer neutral loci. Based on this idea, spreading conservation priorities evenly across the geographic distribution of an island network has been found to capture more neutral genetic variation ([@r200]). However this remains untested in spatially contiguous systems where connectivity is complicated by many additional factors (e.g. land use; [@r497])---the typical situation in most conservation planning exercises.

@@

<!-- summary of methods -->
We detected putatively adaptive genetic variation in `r as.character(as.english(length(env.surrogacy.LST)))` of the 27 plant species. We used two outlier detection methods---that solely used genetic data and did not use environmental data---to identify loci under selection. These methods returned reasonably consistent results (mean `r round(mean(spp.outlier.methods.congruence, na.rm=TRUE),2)` % loci per species assigned the same classification $\pm$ `r round(sd(spp.outlier.methods.congruence, na.rm=TRUE),2)` SD; Dataset S1). Of the species that were associated with adaptive genetic variation, only a small proportion of loci were classified as being adaptive (mean `r round(mean(spp.outlier.loci.props, na.rm=TRUE),2)` % loci per species $\pm$ `r round(sd(spp.outlier.loci.props, na.rm=TRUE),2)` SD). After identifying the loci showing strong signals of selection, we used non-metric multi-dimensional scaling (NMDS) analyses to identify the main gradients of adaptive (if detected) and neutral variation for each species. Generally, only a small number of continuous dimensions were needed to sufficiently describe their patterns of adaptive ($K = `r unique(na.omit(adaptive.k))`$ for all species; Table S1) and neutral genetic variation ($`r min(neutral.k)` \leq K \leq `r max(neutral.k)`$ for all species; Table S1). The resulting ordinations were used to construct an adaptive (if detected) and neutral genetic space for each species. The spatial distribution of these genetic spaces (Figs. S1--S27) generally showed substantial spatial auto-correlation (e.g. _Cerastium uniflorum_ and _Dryas octopetala_, Figs. S6 and S8), suggesting that the proposed surrogate variables have the potential to be effective for conservation planning.

<!-- correlation analysis -->
We verified that the spatial patterns in environmental variation correlated with the broad-scale patterns in adaptive genetic variation for each species. We also verified that there was a correlation between neutral genetic variation and variation in geographic position. For each species, we constructed dissimilarity matrices expressing differences between the planning units where individuals were detected based on the units' (i) environmental characteristics, (ii) geographic position, and also the (iii) adaptive (if adaptive loci detected), and (iv) neutral genetic characteristics of the individuals found inside them. The spatial patterns of environmental variation were significantly correlated with the patterns of adaptive genetic variation for `r as.character(as.english(sum(correlation.model.summary.DF$Environmental.corr.P<0.05,na.rm=TRUE)))` of the `r as.character(as.english(sum(!is.na(correlation.model.summary.DF$Environmental.corr.P))))` species associated with adaptive variation (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Environmental.corr.P<0.05)$Environmental.Marginal.R2),2)` SD for significant models; Table S2). Similarly, geographic distance between planning units was also correlated significantly with the spatial patterns of neutral genetic variation among planning units for `r sum(correlation.model.summary.DF$Geographic.corr.P<0.05,na.rm=TRUE)` of the `r sum(!is.na(correlation.model.summary.DF$Geographic.corr.P))` species (_P_ < 0.05; mean `r round(mean(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` marginal $R^2$ $\pm$ `r round(sd(filter(correlation.model.summary.DF, Geographic.corr.P<0.05)$Geographic.Marginal.R2),2)` SD for significant models; Table S2). Thus, for most species, planning units that contained different environmental conditions contained individuals with different adaptive genetic characteristics, and planning units that were further apart tended to contain individuals with different neutral genetic characteristics. After verifying these correlations, our next step was to determine whether they were strong enough to be useful for conservation planning.

<!-- surrogacy analysis -->
For each species, we generated a suite of `r format(raptr.params.LST[[MODE]]$surrogacy.analysis$random.replicates, big.mark=',', scientific=FALSE)` random prioritizations, and calculated the proportion of environmental variation and adaptive (if present) genetic variation that each prioritization captured. We then repeated this process, and calculated the proportion of variation in geographic position and neutral genetic variation they sampled. The environmental and geographic variables were moderately effective predictors for the genetic variation represented by the randomly generated prioritizations (Figs. 1, S28--S29). The relationship between the proportion of genetic and surrogate variation secured in the prioritizations varied among the species (species $\times$ proportion of environmental variation interaction term: ${\chi^2}_{`r env.surrogacy.AOV$df[[4]]`} = {`r round(env.surrogacy.AOV$Resid..Dev[[4]],2)`}; P = `r pvalString(env.surrogacy.AOV$Pr..Chi.[[4]])`$; species $\times$ proportion of geographic variation interaction term: ${\chi^2}_{`r round(geo.surrogacy.AOV$df[[4]],2)`} = `r round(geo.surrogacy.AOV$Resid..Dev[[4]],2)`$; $P$ `r pvalString(geo.surrogacy.AOV$Pr..Chi.[[4]])`). _Post-hoc_ analyses showed that these relationships were positive for all species (environmental: minimum \textit{Z} = `r round(max(env.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; geographic: minimum \textit{Z} = `r round(min(geo.surrogate.spp.DF$statistic),2)`, maximum $P `r pvalString(max(env.surrogate.spp.DF$p.value))`$; Table S3). After establishing that the proportion of surrogate variation secured in a prioritization also predicted the proportion of genetic variation it secured, our final step was to determine if prioritizations generated using targets to represent variation in the surrogate variables were more effective at representing intra-specific genetic variation.

```{r figure-1, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4.5, out.width='3.42in', out.height='2.5in', fig.cap=paste0('The proportion of adaptive (a; ',sum(!is.na(correlation.model.summary.DF$Environmental.corr.P)),' species) and neutral (b; ',sum(!is.na(correlation.model.summary.DF$Geographic.corr.P)),' species) genetic variation represented in a suite of randomly generated prioritizations for different species as a function of the surrogate variation they also secured (lines). Panels (c) and (d) show the distribution of $R^2$ values for the ',sum(!is.na(correlation.model.summary.DF$Environmental.corr.P)) + sum(!is.na(correlation.model.summary.DF$Geographic.corr.P)),' models shown in panels (a) and (b).')}
# make figure
p1 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=env.surrogacy.DF) +
  xlab('Environmental variation\nsecured (%)') +
  ylab('Adaptive genetic\nvariation secured (%)') +
  theme_classic() + theme(axis.line.x=element_line(color='black'),
    axis.line.y=element_line(color='black')) +
  coord_cartesian(xlim=c(0, 1)) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_y_continuous(limits=c(0,1), labels=function(x) {x*100})
for (i in seq_along(env.surrogacy.LST))
  p1 <- p1 + stat_smooth(method='glm', method.args=list(family='binomial'),
    data=filter(env.surrogacy.DF, Species==unique(Species)[i]),
    fullrange=TRUE, se=FALSE, colour='black')
p1 <- p1 + geom_text(x=-Inf, y=Inf, hjust=-0.25, vjust=1.5, label='(a)')

p2 <- ggplot(aes(x=surrogate.held,y=genetic.held), data=geo.surrogacy.DF) +
  xlab('Geographic variation\nsecured (%)') +
  ylab('Neutral genetic\nvariation secured (%)') +
  theme_classic() + theme(axis.line.x=element_line(color='black'),
    axis.line.y=element_line(color='black', colour='transparent')) +
  coord_cartesian(xlim=c(0, 1)) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_x_continuous(limits=c(0,1), labels=function(x) {x*100}) +
  scale_y_continuous(limits=c(0,1), labels=function(x) {x*100})
for (i in seq_along(geo.surrogacy.LST))
  p2 <- p2 + stat_smooth(method='glm', method.args=list(family='binomial'),
    data=filter(geo.surrogacy.DF, Species==unique(Species)[i]),
    fullrange=TRUE, se=FALSE, colour='black')
p2 <- p2 + geom_text(x=-Inf, y=Inf, hjust=-0.25, vjust=1.5, label='(b)')

@@

<!-- scenario analysis -->
To determine whether environmental and geographic targets could improve the effectiveness of prioritizations in representing genetic variation, we generated prioritizations using (i) "amount targets" reflecting the traditional approach of representing a certain proportion of each species' geographic distribution, (ii) "amount and surrogate targets" where we targeted representation of the environmental and geographic surrogates as well as a certain proportion of each species' geographic distribution, and (iii) "amount and genetic targets" where we targeted representation of the directly measured genetic variation as well as a certain proportion of each species' geographic distribution. To determine if these patterns were robust, for each of the three combinations of the targets, we generated prioritizations under four scenarios: (i) single-species with equal costs, (ii) single species with acquisition costs, (iii) multi-species with equal costs, and (iv) multi-species with acquisition costs (Figs. 2 and S30).

```{r figure-2, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=(5*0.9), fig.height=(7.1*0.9), out.width='3.42in', out.height='4in', fig.cap='The proportion of adaptive and neutral genetic variation secured for each species in solutions based on different targets and scenarios. Box plots show the median, and the 25th and 75th percentiles. Whiskers show the data at 1.5 times the inter-quantile range. Points show data outside the whiskers. Red lines show the target proportion of genetic variation.'}
# prepare data for plotting
scenario.SDF <- scenario.DF %>%
  mutate(Context = gsub('opportunity', 'acquisition', Context, fixed = TRUE)) %>%
  group_by(Prioritisation, Context, Metric, Species) %>%
  summarize(genetic.held=mean(genetic.held, na.rm=TRUE)) %>%
  data.frame()

@@

The proportion of genetic variation secured in a given prioritization was not found to depend on any interaction between the planning scenario or targets used to generate the prioritization, or the type of genetic variation measured (${\chi^2}_{`r scenario.ANOVA.1.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.1.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.1.DF$p.value[[2]])`$). The proportion of genetic variation secured in a prioritization did depend, however, on the targets used to generate the prioritization (${\chi^2}_{`r scenario.ANOVA.4.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.4.DF$statistic[[2]],2)`$; $P `r pvalString(scenario.ANOVA.4.DF$p.value[[2]])`$). Most notably, prioritizations based on the surrogate variables represented significantly more genetic variation than amount-based prioritizations (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & surrogate\ntargets')$genetic.held*100),2)` SD overall genetic variation secured versus `r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount targets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[1]],2)`$; $P `r pvalString(scenario.GLHT.2.DF$p.value[[1]])`$), and were not distinguishable from prioritizations based on measured genetic data (`r round(mean(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Prioritisation=='Amount & genetic\ntargets')$genetic.held*100),2)` SD; $Z_1 = `r round(scenario.GLHT.2.DF$statistic[[3]],2)`$; $P `r pvalString(scenario.GLHT.2.DF$p.value[[3]])`$). These results indicate that environmental and geographic surrogates in this case far outperform traditional amount-based conservation planning, and perform almost as well as a conservation plan based directly on genetic data.

Overall, the prioritizations tended to secure a greater proportion of adaptive genetic variation than neutral genetic variation (adaptive: mean `r round(mean(filter(scenario.sub.DF,Metric=='Adaptive variation')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Metric=='Adaptive variation')$genetic.held*100),2)` secured across all prioritizations, neutral: mean `r round(mean(filter(scenario.sub.DF,Metric=='Neutral variation')$genetic.held*100),2)` % $\pm$ `r round(sd(filter(scenario.sub.DF,Metric=='Neutral variation')$genetic.held*100),2)`; ${\chi^2}_{`r scenario.ANOVA.3.DF$Chi.Df[[2]]`} = `r round(scenario.ANOVA.3.DF$statistic[[2]],2)`$; _P_ `r pvalString(scenario.ANOVA.3.DF$p.value[[2]])`). Thus, regardless of the targets used to generate a prioritization, or the planning context under which the prioritization was generated, prioritizations tend to secure more adaptive than neutral genetic genetic variation.
